Stochastic Approximation Monte Carlo with a Dynamic Update Factor (SAD)

### 2018-12-12 papers/histogram/sad.tex
made remaining changes to the SAD paper

\subsection{SAMC}
The Stochastic Approximation Monte Carlo (SAMC) algorithm addresses
the lack of convergence of Wang-Landau's approach with a simple
schedule by which the update factor $\gamma$ is continuously
decreased~\cite{liang2007stochastic, werlich2015stochastic,
  schneider2017convergence}.  The update factor is defined in the
original implementation~\cite{liang2007stochastic} in terms of an
arbitrary tunable parameter $t_0$.
\begin{align}
\gamma_{t}^{\text{SA}} =\frac{t_0}{\max(t_0,t)}
\end{align}
where as above $t$ is the number of moves that have been attempted.
The major advantage that SAMC offers is its proven convergence.
Provided the update factor satisfies
\begin{align}
\sum_{t=1}^\infty \gamma_{t} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty \gamma_{t}^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$, Liang has shown that the weights are proven
to converge to the true density of states~\cite{liang2006theory,
  liang2007stochastic}.  In addition, the energy range need not be
known a priori.  The time to converge depends only on the choice of
parameters $t_0$.  Unfortunately, there is no
prescription for finding an acceptable value for $t_0$, and
while the algorithm formally converges, for a poor choice of $t_0$
that convergence can be far too slow to be practical.
Liang et al. give a rule of thumb in which $t_0$
is chosen in the range from $2N_S$ to $100N_S$ where $N_S$ is the number
of energy bins~\cite{liang2007stochastic}.  Schneider et al. found
that for the
Ising model this heuristic is helpful for small spin
systems, but that larger systems require an even higher $t_0$
value~\cite{schneider2017convergence}.  We find that for the square-well
fluid also $t_0$ in some cases needs to be two orders of magnitude higher
than the rule of thumb of $100N_S$.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/N50-lndos-comparison}
  \caption{The entropy of a square well fluid with 50 atoms and filling fraction
        0.3 as the green line.  The green hatched area reflects the
        minimum entropy change needed to converge to the true value.
        The light blue area is the quadratic approximation
        for the change in entropy.  The vertical dotted lines represent
        the energy corresponding to $T=0.2$ and $T=\infty$.}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/gamma-n500}
  \caption{The update factor $\gamma_t$ for $N=500$ versus iteration for three
    different methods: WL, SAMC, and SAD.
    method.}\label{fig:gamma-vs-t}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/gamma-n50}
  \caption{The update factor $\gamma_t$ for $N=50$ versus iteration for three
    different methods: WL, SAMC, and SAD.}
\end{figure}

@@

For the square-well fluid, our simulations treat systems with a
well-width of $\lambda = 1.3$ and a filling fraction of
$\eta = 0.3$. We also explore various translation scales $\delta_0 = 0.005\sigma$ and
$0.05\sigma$. The simulations explore the
energy space of the systems at reduced temperatures of
$T_{\text{min}} = 0.2$ and $0.5$.  All simulations lead to the minimum important
energy $E_{\min}$ and maximum entropy energy $E_{\max}$
being calculated (with the exception of the WL methods where both
of these parameters are needed a priori).  In the sections below, we
examine the systems parameters impact on the convergence of the methods.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error-default.pdf}
  \caption{The average entropy error for each MC method is shown for $\eta = 0.3$ and $T_{\min} = 1/3$ with
  SAMC $t_0 = 10^5$ serving as the reference.}\label{fig:N50-ff0.3-avg-error}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error-slow.pdf}
  \caption{The average entropy error for each MC method is shown for $\eta = 0.3$ and $T_{\min} = 1/3$ with
  SAMC $t_0 = 10^5$ from Fig.~\ref{fig:N50-ff0.3-avg-error}
  serving as the reference.}\label{fig:N50-ff0.3-avg-error-slow}
\end{figure}
%\begin{figure}
  %\includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error-fast.pdf}
  %\caption{The average entropy error for each method is shown for $\eta = 0.3$ with
  %SAMC followed by TMMC serving as the reference system.}\label{fig:N50-ff0.3-avg-error-fast}
%\end{figure}

@@

Figure~\ref{fig:N50-ff0.3-avg-error} shows the average entropy vs moves
for a system consisting of 50 atoms that are placed in a periodic
simple cubic unit cell with lattice constant chosen to give a filling
fraction of $\eta = 0.3$.  These particles are randomly inserted into
the system and random moves are proposed (with a translation scale
$\delta_0 = 0.05\sigma$). For the $\eta = 30\%$ and $N = 50$ system, we
explore the choice of $\delta_0$ as a parameter.

@@

Smaller translation
scales cause the given method to take additional time to explore all
energies. Ideal methods should scale linearly with the $\delta_0$, that
is, a decrease in the translation scale should result in a proportional
increase in convergence time. However, as can be seen from
Fig.~\ref{fig:N50-ff0.3-avg-error-slow} SAMC methods that rapidly
converged for $\delta = 0.05\sigma$ do not proportionaly converge for a
translation scale of $\delta = 0.005\sigma$. We see that SAD handles
arbitrary translation scales and proportionaly converges based on the
translation scale that is chosen.

\begin{figure}
\includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_50-ff0_17-N256-entropy-error-defualt.pdf}
  \caption{The average entropy error for each MC method is shown for $\eta = 0.17$ and $T_{\min} = 1$ with
  SAD serving as the reference.}\label{fig:n256}
\end{figure}
\begin{figure}
\includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N500-entropy-error-default.pdf}
  \caption{The average entropy error for each MC method is shown for $\eta = 0.3$ and $T_{\min} = 1/2$ with
  SAMC $t_0 = 10^6$ serving as the reference.}\label{fig:n500-ff0.3}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/N500-lndos-comparison}
  \caption{Use this to create description of the system.}\label{fig:n500-lndos-comparison}
\end{figure}

\subsection{A periodic system with 500 atoms}

A system is configured with dimensions $x = y = z = 19\sigma$.  A
filling fraction $\eta = 0.3$ is chosen which yields $N = 500$
particles in a cubical lattice system.  These particles are randomly
inserted into the system and random moves are proposed (with a
translation scale $\delta_0 = 0.05\sigma$).

{\color{red}
$S_{\max} = 1206$ and $S_{\min} = 2606$ for this system.  The number of
important energy states for this system is then $N_S = S_{\min} -
S_{\max} = 1400$. The reduced temperature of $1/2$ is chosen for the
larger system in order to examine the convergence properties of SAD in
a reasonable simulation time.
}

For the larger system, SAMC parameter values of $t_0$ that gave rapid
convergence for the $N = 50$ system with $\delta = 0.05\sigma$ do not
converge rapidly.  Several additional larger $t_0$ parameters are
chosen to compare with SAD.  SAD has significantly smaller fluctuations
in the average entropy error for a given number of moves.

{\color{red}
Figure~\ref{fig:n500-lndos-comparison} shows the calculated entropy for
the $N = 500$ system. The grey dash lines indicate the $S_{\max} = 1206$
point and the $S_{\min} = 2606$. Only the reference method is shown.
}

@@

A close examination of Fig.~\ref{fig:gamma-vs-t} shows that $\gamma$
for the Wang-Landau algorithm decreases much more rapidly than
$\nicefrac{1}{\sqrt{t}}$. It can be observed  for the simulations that
were carried out that all the methods from the weight-based family
struggle to converge if the update factor decreases too rapidlly.
Fig.~\ref{fig:n500-ff0.3} shows WL and SAMC with inappropriate choices
of $t_0$ i.e. less than $\nicefrac{10^5}{\sqrt{t}}$ failing to converge
in a reasonable amount of time. Thus, the update factor $\gamma$, shown
as a function of simulation moves in Fig.~\ref{fig:gamma-vs-t}, proves
useful in determining which methods will converge in a reasonable
amount of time. We compare each method to a reference SAMC
simulation for an ideal choice of $t_0$.  This gives an unbiased
comparison with our developed method SAD. The dynamic version of SAMC
which we call SAD performs as well as SAMC (assuming the user can
correctly identify an appropriate choice of $t_0$ for SAMC).  SAD
outperforms WL and WL-TMMC for the same given number of moves. The
absence of user defined parameters makes SAD an exciting alternative to
methods where user experience or successive implementation iteration
are needed to achieve reasonable results.

>>>>

\subsection{SAMC}
The Stochastic Approximation Monte Carlo (SAMC) algorithm addresses
the lack of convergence of Wang-Landau's approach with a simple
schedule by which the update factor $\gamma$ is continuously
decreased~\cite{liang2007stochastic, werlich2015stochastic,
  schneider2017convergence}.  The update factor is defined in the
original implementation~\cite{liang2007stochastic} in terms of an
arbitrary tunable parameter $t_0$.
\begin{align}
\gamma_{t}^{\text{SA}} =\frac{t_0}{\max(t_0,t)}
\end{align}
where as above $t$ is the number of moves that have been attempted.
The major advantage that SAMC offers is its proven convergence.
Provided the update factor satisfies
\begin{align}
\sum_{t=1}^\infty \gamma_{t} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty \gamma_{t}^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$, Liang has shown that the weights are proven
to converge to the true density of states~\cite{liang2006theory,
  liang2007stochastic}.  In addition, the energy range need not be
known a priori.  The time to converge depends only on the choice of
parameters $t_0$.  Unfortunately, there is no
prescription for finding an acceptable value for $t_0$, and
while the algorithm formally converges, for a poor choice of $t_0$
that convergence can be far too slow to be practical.
Liang et al. give a rule of thumb in which $t_0$
is chosen in the range from $2N_S$ to $100N_S$ where $N_S$ is the number
of energy bins~\cite{liang2007stochastic}.  Schneider et al. found
that for the
Ising model this heuristic is helpful for small spin
systems, but that larger systems require an even higher $t_0$
value~\cite{schneider2017convergence}.  Also, we find that for the square-well
fluid $t_0$ in some cases needs to be two orders of magnitude higher
than the rule of thumb of $100N_S$.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/N50-lndos-comparison}
  \caption{The entropy of a square well fluid with 50 atoms and filling fraction
        0.3 as the green line.  The green hatched area reflects the
        minimum entropy change needed to converge to the true value.
        The light blue area is the quadratic approximation
        for the change in entropy.  The vertical dotted lines represent
        the energy corresponding to $T=0.\overline{3}$ and $T=\infty$.}
\end{figure}

@@

%\begin{figure}
  %\includegraphics[width=\columnwidth]{figs/gamma-n500}
  %\caption{The update factor $\gamma_t$ for $N=500$ versus iteration for three
    %different methods: WL, SAMC, and SAD.
    %method.}\label{fig:gamma-vs-t}
%\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/gamma-n256}
  \caption{The update factor $\gamma_t$ for $N=256$ versus iteration for three
    different methods: WL, SAMC, and SAD.
    method.}\label{fig:gamma-vs-t}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/gamma-n50}
  \caption{The update factor $\gamma_t$ for $N=50$ versus iteration for three
    different methods: WL, SAMC, and SAD.}
\end{figure}

@@

For the square-well fluid, our simulations treat systems with a
well-width of $\lambda = 1.3$ and $\lambda = 1.5$ and a filling fraction of
$\eta = 0.3$ and $\eta = 0.17$. We also explore various translation scales $\delta_0 = 0.005\sigma$ and
$0.05\sigma$. The simulations explore the
energy space of the systems at reduced temperatures of
$T_{\text{min}} = 0.\overline{3}$ and $1.0$.  All simulations lead to the minimum important
energy $E_{\min}$ and maximum entropy energy $E_{\max}$
being calculated (with the exception of the WL methods where both
of these parameters are needed a priori).  In the sections below, we
examine the systems parameters impact on the convergence of the methods.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error-default.pdf}
  \caption{The average entropy error for each MC method is shown for $\eta = 0.3$ and $T_{\min} = 1/3$ with
  SAMC $t_0 = 10^5$ serving as the reference.}\label{fig:N50-ff0.3-avg-error}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error-slow.pdf}
  \caption{The average entropy error for each MC method is shown for $\eta = 0.3$ and $T_{\min} = 1/3$ with
  SAMC $t_0 = 10^5$ from Fig.~\ref{fig:N50-ff0.3-avg-error}
  serving as the reference.}\label{fig:N50-ff0.3-avg-error-slow}
\end{figure}

@@

Figures~\ref{fig:N50-ff0.3-avg-error}
and~\ref{fig:N50-ff0.3-avg-error-slow} show the average entropy vs
moves for a system consisting of 50 atoms that are placed in a periodic
simple cubic unit cell with lattice constant chosen to give a filling
fraction of $\eta = 0.3$.  These particles are randomly inserted into
the system and random moves are proposed (with a translation scale
$\delta_0 = 0.05\sigma$ and $\delta_0 = 0.005\sigma$ respectively). For
the $\eta = 30\%$ and $N = 50$ system, we explore the choice of
$\delta_0$ as a parameter.

@@

Smaller translation
scales cause the given method to take additional time to explore all
energies. Ideal methods should scale linearly with the $\delta_0$, that
is, a decrease in the translation scale should result in a proportional
increase in convergence time. As can be seen from
Fig.~\ref{fig:N50-ff0.3-avg-error-slow}, SAMC methods that rapidly
converged for $\delta = 0.05\sigma$ do not proportionaly converge for a
translation scale of $\delta = 0.005\sigma$. We see that SAD handles
arbitrary translation scales and proportionaly converges based on the
translation scale that is chosen.

\begin{figure}
\includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_50-ff0_17-N256-entropy-error-default.pdf}
  \caption{The average entropy error for each MC method is shown for $\eta = 0.17$ and $T_{\min} = 1$ with
  SAD serving as the reference.}\label{fig:n256}
\end{figure}
%\begin{figure}
%\includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N500-entropy-error-default.pdf}
  %\caption{The average entropy error for each MC method is shown for $\eta = 0.3$ and $T_{\min} = 1/2$ with
  %SAMC $t_0 = 10^6$ serving as the reference.}\label{fig:n500-ff0.3}
%\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/N256-lndos-comparison}
  \caption{Use this to create description of the system.}\label{fig:n256-lndos-comparison}
\end{figure}

\subsection{A periodic system with 256 atoms}

Figure~\ref{fig:n256} shows the average entropy vs moves for a system
consisting of 256 atoms that are placed in a periodic simple cubic unit
cell with lattice constant chosen to give a filling fraction of $\eta =
0.17$.   These particles are randomly inserted into the system and
random moves are proposed (with a translation scale $\delta_0 =
0.05\sigma$).

{\color{red}
$S_{\max} = 509$ and $S_{\min} = 915$ for this system.  The number of
important energy states for this system is then $N_S = S_{\min} -
S_{\max} = 406$. The reduced temperature of $1.0$ and well-width of $\lambda = 1.5$
are chosen in order to ensure that the simulation encounters liquid vapor coexistence
and that the results are comparable with NIST.
}

For the larger system, SAMC parameter values of $t_0$ that gave rapid
convergence for the $N = 50$ system with $\delta = 0.05\sigma$ do not
converge rapidly.  Larger $t_0$ parameters are simulated for
comparison with SAD.  SAD has significantly smaller fluctuations
in the average entropy error for a given number of moves.

{\color{red}
Figure~\ref{fig:n256-lndos-comparison} shows the calculated entropy for
the $N = 256$ system. The grey dash lines indicate the $S_{\max} = 509$
point and the $S_{\min} = 915$. Only the reference method is shown.
}

@@

A close examination of Fig.~\ref{fig:gamma-vs-t} shows that $\gamma$
for the Wang-Landau algorithm decreases much more rapidly than
$\nicefrac{1}{\sqrt{t}}$. It can be observed  for the simulations that
were carried out that all the methods from the weight-based family
struggle to converge if the update factor decreases too rapidlly.
Fig.~\ref{fig:N50-ff0.3-avg-error-slow} shows SAMC with inappropriate choices
of $t_0$ i.e. less than $\nicefrac{10^5}{\sqrt{t}}$ failing to converge
in a reasonable amount of time. Thus, the update factor $\gamma$, shown
as a function of simulation moves in Fig.~\ref{fig:gamma-vs-t}, proves
useful in determining which methods will converge in a reasonable
amount of time. We compare each method to a reference SAMC
simulation for an ideal choice of $t_0$.  This gives an unbiased
comparison with our developed method SAD. The dynamic version of SAMC
which we call SAD performs as well as SAMC (assuming the user can
correctly identify an appropriate choice of $t_0$ for SAMC).  SAD
outperforms WL and WL-TMMC for the same given number of moves. The
absence of user defined parameters makes SAD an exciting alternative to
methods where user experience or successive implementation iteration
are needed to achieve reasonable results.
